The function--sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-08-18 dictionary.tex
Changed the language on bounded and finite quasicharacter sheaves
Made a home for applications to Abelian varieties

\begin{proposition}\label{prop:mor_etale}
Let $G$ be an etale commutative group scheme over $\Fq$.
If $\qcs{L}$ and $\qcs{L}'$ are quasicharacter sheaves on $G$ then
every $\rho\in \Hom(\qcs{L},\qcs{L}')$ is either trivial or an isomorphism. Moreover, the trace map induces an isomorphism of groups
\[
\Aut(\qcs{L}) \to \Hom(G(\bFq)_{\Weil{}}, \EEx).
\]
\end{proposition}

@@

Note that Proposition~\ref{prop:mor_etale} tells us that if $G$ is an etale commutative group scheme over $\Fq$ then every hom-set is either trivial or $\left(G(\bFq)_{\Weil{}}\right)^*$, non-canonically.

\section{Main results}\label{sec:main}

@@

We will identify $\bQC(G)$ with its image under $B_G$ and say that a quasicharacter sheaf $\qcs{L} \in \QC(G)$ \emph{descend to $G$} if it is isomorphic to some $B_G(\qcs{E}, \mu_\qcs{E})$.
%We will see in Corollary~\ref{cor:bounded-and-finite} that bounded quasicharacter sheaves are exactly quasicharacter sheaves with bounded Trace of Frobenius.\todo{Check this, later.}

@@

\begin{lemma}\label{lem:bounded_connected}
If $G$ is a connected commutative algebraic group over $\Fq$ then every quasisharacter sheaf on $G$ is finite and $B_G : \bQC(G) \to \QC(G)$ is an equivalence of categories.
\end{lemma}

@@

\begin{proof}
In Lemma~\ref{lem:bounded_connected} we saw that every quasicharacter sheaf $\qcs{L}$ on $G$ descends to bounded quasicharacter sheaf $\qcs{E}$ on $G$. Since the functor $B_G : \bQC(G) \to \QC(G)$ is full and faithful, $\Aut(\qcs{L}) = \Aut(\qcs{E})$.
From here, Deligne's function--sheaf dictionary for connected commutative algebraic groups over finite fields \citelist{\cite{deligne:SGA4.5}*{Sommes trig.}, \cite{laumon:87a}*{1.1.3}}  gives us all we need, as we briefly recall.
The key observation is that, by etale descent, pull-back by the Lang isogeny $\Lang : G\to G$ defines an equivalence of categories from $\bQC(G)$ to the category of $1$-dimensional $\ell$-adic representations $V$ of $G(\Fq)$ equipped with an isomorphism $V \to V\otimes V$ such that the trace of $\Lang^*\qcs{E}$ is $\trFrob{\qcs{E}}^{-1}$ for every bounded quasicharacter sheaf $\qcs{E}$ on $G$. 
Note that since $V$ is $1$-dimensional, the choice of $V \to V\otimes V$ is exactly the choice of an isomorphism $V\iso \EE$.
Using the equivalence given by $\Lang^*$, if $\Lang^*\qcs{E} = V$ with character $\psi$ and isomorphism $V\to V\otimes V$ then 
$\qcs{E} \iso (\Lang_* V)_\psi$, the $\psi$-isotypic component of the push-forward $\Lang_* V$ of the constand sheaf $V$ on $G$, equipped with an isomorphism $\mu_\qcs{E} : m^* \qcs{E} \to \qcs{E} \boxtimes\qcs{E}$ determined by $V\to V\otimes V$.
Moreover, $\Aut(\qcs{E})$ is the group of automorphisms of $\rho : V\to V$ for which 
\[
\begin{tikzcd}
\arrow{d}{} V \arrow{r}{\rho} & V\arrow{d}{}\\
V\otimes V \arrow{r}{\rho\otimes \rho} & V\otimes V
\end{tikzcd}
\]
commutes. 
Since the only such isomorphism $\rho$ is $\id_V$, it follows that $\Aut(\qcs{E}) = 1$, completing the proof.
For the last part of the proposition, argue as in the analogous part of Proposition~\ref{prop:mor_etale}.
\end{proof}

@@

\subsection{Finite quasicharacter sheaves}

@@

\begin{proposition}\label{prop:finite}
Let $f: H \to G$ be a discrete isogeny; let $A$ be the kernel of $f$.
Let $\psi : A \to \EEx$ be a character.
Then $(f_* \EE)_\psi$ (the $\psi$-isotypic component of $f_*\EE$) is a bounded quasicharacter on $G$.
\end{proposition}

@@

We say that a morphism $f : H\to G$ of commutative group $\Fq$-schemes is a discrete isogeny if it is a finite, surjective \'etale morphism and
the action of $\Gal(\bFq/\Fq)$ on the \'etale group scheme $\ker f$ is trivial.

\begin{lemma}\label{lemma:ext}
Every discrete isogeny onto $G^0$ extends to a discrete
isogeny onto $G$ inducing an isomorphism on component groups.
\end{lemma}

@@

\subsection{The function--sheaf dictionary for smooth, commutative group schemes}
\label{sec:snake}

@@

\begin{theorem}\label{thm:hom}
Let $G$ be a smooth, commutative group scheme over $\Fq$.
For all quasicharacter sheaves $\qcs{L}$ and $\qcs{L}'$ on $G$,
every $\rho \in \Hom(\qcs{L},\qcs{L}')$ is either trivial or an isomorphism.
Moreover, $\Aut(\qcs{L}) \iso \Hom(\pi_0(\bG)_{\Weil{}},\EEx)$, canonically.
\end{theorem}
\begin{proof} 
For the first part of the proposition, argue as in the analogous part of Proposition~\ref{prop:mor_etale}.
For the second, fix $\qcs{L} = (\gqcs{L},\mu,\phi)$ and consider the homomorphism $\Aut(\qcs{L}) \to \Hom(\bG_{\Weil{}},\EEx)$ defined in Proposition~\ref{prop:mor_etale}.
This group homomorphism is clearly injective: if $\rho_x = \id_{\qcs{L}_x}$ for every $x\in \bG$ then $\rho = \id_{\qcs{L}}$.
Note that the image of $\Aut(\qcs{L}) \to \Hom(\bG_{\Weil{}},\EEx)$ is contained in $\Hom_\text{cts}(\bG_{\Weil{}},\EEx)$, where $\bG_{\Weil{}}$ is given the topology of the base of the espace étalé attached to $\gqcs{L}$.
Since $\ell$ is invertible in $\Fq$, it follows that $\Hom_\text{cts}(\bG_{\Weil{}},\EEx) = \Hom(\pi_0(\bG_{\Weil{}}),\EEx)$. 
Note that $\pi_0(\bG_{\Weil{}})=\pi_0(\bG)_{\Weil{}}$. 
To see that $\Aut(\qcs{L}) \to \Hom(\pi_0(\bG)_{\Weil{}},\EEx)$ is surjective, begin with $\theta\in\Hom(\pi_0(\bG)_{\Weil{}},\EEx)$ and, for each $[x] \in \pi_0(\bG)_{\Weil{}}$ define $\bar\rho_y: \gqcs{L}_y \to \gqcs{L}_y$ by scalar multiplication by $\theta([x])\in \EEx$ for each $y\in [x]$.
This defines an isomorphism $\bar\rho : \gqcs{L}\to \gqcs{L}$ of local systems on $\bG$ compatible with $\mu$ and $\phi$, and thus an isomorphism $\rho :\qcs{L}\to \qcs{L}$ with maps to $\theta$ under $\Aut(\qcs{L}) \to \Hom(\pi_0(\bG)_{\Weil{}},\EEx)$.
\end{proof}

\section{Application to \texorpdfstring{$p$}{p}-adic tori}

@@

@@

\subsection{Recovery}

We are in a good position to understand $\QC(G)$ when $G$ is a connected algebraic group over $\Fq$, since we can now describe objects and maps in this category in very concrete terms. We begin by recovering a classical result.

Morphisms in $\QC(G)$ are also very simple.

\begin{proposition}
If $G$ is a connected algebraic group then $\Aut_{\QC(G)}(\qcs{L})$ is . . . .
\end{proposition}

\begin{proof}
Pick up from the proof of Proposition~\ref{prop:autornaught}.
\end{proof}

>>>>

\begin{proposition}\label{prop:autornaught_etale}
Let $G$ be an etale commutative group scheme over $\Fq$.
If $\qcs{L}$ and $\qcs{L}'$ are quasicharacter sheaves on $G$ then
every $\rho\in \Hom(\qcs{L},\qcs{L}')$ is either trivial or an isomorphism. Moreover, the trace map induces an isomorphism of groups
\[
\Aut(\qcs{L}) \to \Hom(G(\bFq)_{\Weil{}}, \EEx).
\]
\end{proposition}

@@

Note that Proposition~\ref{prop:autornaught_etale} tells us that if $G$ is an etale commutative group scheme over $\Fq$ then every hom-set is either trivial or $\left(G(\bFq)_{\Weil{}}\right)^*$, non-canonically.

\section{The function--sheaf dictionary for smooth, commutative group schemes}\label{sec:main}

@@

We will identify $\bQC(G)$ with its image under $B_G$ and say that a quasicharacter sheaf $\qcs{L} \in \QC(G)$ \emph{descends to $G$} if it is isomorphic to some $B_G(\qcs{E}, \mu_\qcs{E})$.
%We will see in Corollary~\ref{cor:bounded-and-finite} that bounded quasicharacter sheaves are exactly quasicharacter sheaves with bounded Trace of Frobenius.\todo{Check this, later.}

@@

\begin{lemma}\label{lem:bounded_connected}
If $G$ is a connected commutative algebraic group over $\Fq$ then $B_G : \bQC(G) \to \QC(G)$ is an equivalence of categories.
\end{lemma}

@@

\begin{proof}
In Lemma~\ref{lem:bounded_connected} we saw that every quasicharacter sheaf $\qcs{L}$ on $G$ descends to $G$; let $\qcs{E}$ be an object in $\bQC(G)$ for which $B_G(\qcs{E}) \iso \qcs{L}$.
 Since the functor $B_G : \bQC(G) \to \QC(G)$ is full and faithful, $\Aut(\qcs{L}) = \Aut(\qcs{E})$.
From here, Deligne's function--sheaf dictionary for connected commutative algebraic groups over finite fields \citelist{\cite{deligne:SGA4.5}*{Sommes trig.}, \cite{laumon:87a}*{1.1.3}}  gives us all we need, as we briefly recall.
The key observation is that, by etale descent, pull-back by the Lang isogeny $\Lang : G\to G$ defines an equivalence of categories from $\bQC(G)$ to the category of $1$-dimensional $\ell$-adic representations $V$ of $G(\Fq)$ equipped with an isomorphism $V \to V\otimes V$ such that the trace of $\Lang^*\qcs{E}$ is $\trFrob{\qcs{E}}^{-1}$ for every $\qcs{E}$ in $\bQC(G)$. 
Note that since $V$ is $1$-dimensional, the choice of $V \to V\otimes V$ is exactly the choice of an isomorphism $V\iso \EE$.
Using the equivalence given by $\Lang^*$, if $\Lang^*\qcs{E} = V$ with character $\psi$ and isomorphism $V\to V\otimes V$ then 
$\qcs{E} \iso (\Lang_* V)_\psi$, the $\psi$-isotypic component of the push-forward $\Lang_* V$ of the constand sheaf $V$ on $G$, equipped with an isomorphism $\mu_\qcs{E} : m^* \qcs{E} \to \qcs{E} \boxtimes\qcs{E}$ determined by $V\to V\otimes V$.
Moreover, $\Aut(\qcs{E})$ is the group of automorphisms of $\rho : V\to V$ for which 
\[
\begin{tikzcd}
\arrow{d}{} V \arrow{r}{\rho} & V\arrow{d}{}\\
V\otimes V \arrow{r}{\rho\otimes \rho} & V\otimes V
\end{tikzcd}
\]
commutes. 
Since the only such isomorphism $\rho$ is $\id_V$, it follows that $\Aut(\qcs{E}) = 1$, completing the proof.
For the last part of the proposition, argue as in the analogous part of Proposition~\ref{prop:autornaught_etale}.
\end{proof}

@@

\subsection{Discrete isogenies}\label{sec:discrete_isogenies}

@@

\begin{proposition}\label{prop:finite}
Let $f: H \to G$ be a discrete isogeny; let $A$ be the kernel of $f$.
Let $\psi : A \to \EEx$ be a character.
Then $(f_* \EE)_\psi$ (the $\psi$-isotypic component of $f_*\EE$) is an object in $\bQC(G)$.
\end{proposition}

@@

We say that a morphism $f : H\to G$ of commutative group $\Fq$-schemes is a {\it discrete isogeny} to $G$ if it is a finite, surjective \'etale morphism and
the action of $\Gal(\bFq/\Fq)$ on the \'etale group scheme $\ker f$ is trivial.

\begin{lemma}\label{lemma:ext}
Every discrete isogeny to $G^0$ extends to a discrete
isogeny to $G$ inducing an isomorphism on component groups.
\end{lemma}

@@

\subsection{The dictionary}
\label{sec:snake}

@@

\begin{theorem}\label{thm:autornaught}
Let $G$ be a smooth, commutative group scheme over $\Fq$.
For all quasicharacter sheaves $\qcs{L}$ and $\qcs{L}'$ on $G$,
every $\rho \in \Hom(\qcs{L},\qcs{L}')$ is either trivial or an isomorphism.
Moreover, $\Aut(\qcs{L}) \iso \Hom(\pi_0(\bG)_{\Weil{}},\EEx)$, canonically.
\end{theorem}
\begin{proof} 
For the first part of the proposition, argue as in the analogous part of Proposition~\ref{prop:autornaught_etale}.
For the second, fix $\qcs{L} = (\gqcs{L},\mu,\phi)$ and consider the homomorphism $\Aut(\qcs{L}) \to \Hom(\bG_{\Weil{}},\EEx)$ defined in Proposition~\ref{prop:autornaught_etale}.
This group homomorphism is clearly injective: if $\rho_x = \id_{\qcs{L}_x}$ for every $x\in \bG$ then $\rho = \id_{\qcs{L}}$.
Note that the image of $\Aut(\qcs{L}) \to \Hom(\bG_{\Weil{}},\EEx)$ is contained in $\Hom_\text{cts}(\bG_{\Weil{}},\EEx)$, where $\bG_{\Weil{}}$ is given the topology of the base of the espace étalé attached to $\gqcs{L}$.
Since $\ell$ is invertible in $\Fq$, it follows that $\Hom_\text{cts}(\bG_{\Weil{}},\EEx) = \Hom(\pi_0(\bG_{\Weil{}}),\EEx)$. 
Note that $\pi_0(\bG_{\Weil{}})=\pi_0(\bG)_{\Weil{}}$. 
To see that $\Aut(\qcs{L}) \to \Hom(\pi_0(\bG)_{\Weil{}},\EEx)$ is surjective, begin with $\theta\in\Hom(\pi_0(\bG)_{\Weil{}},\EEx)$ and, for each $[x] \in \pi_0(\bG)_{\Weil{}}$ define $\bar\rho_y: \gqcs{L}_y \to \gqcs{L}_y$ by scalar multiplication by $\theta([x])\in \EEx$ for each $y\in [x]$.
This defines an isomorphism $\bar\rho : \gqcs{L}\to \gqcs{L}$ of local systems on $\bG$ compatible with $\mu$ and $\phi$, and thus an isomorphism $\rho :\qcs{L}\to \qcs{L}$ with maps to $\theta$ under $\Aut(\qcs{L}) \to \Hom(\pi_0(\bG)_{\Weil{}},\EEx)$.
\end{proof}

@@

\subsection{Finite and bounded - fragments}

\section{Applications to algebraic tori and Abelian varieties over \texorpdfstring{$p$}{p}-adic fields}

@@
